---
title: "Mean Expression and Dropout Rate have Consistent Relationship across Cell Types and Data Sets"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 6, fig.height = 4,
                      comment = NA)
library(Matrix)
library(gplots)
library(ggplot2)
library(reshape2)
library(dplyr)
library(broom)
library(ggpubr)
library(data.table)
library(cramer)
get_positive_mean = function(x){
  ind = which( x > 0)
  if(length(ind) > 0){
    mean(x[ind])
  }else{
    return(0)
  }
}
preprocess = function(X, label, upper, normalize){
  if(normalize){
    X = X/rowSums(X) * 3000
  }
  Y = X
  X = X>0 * 1

  numcounts = colSums(X)
  ind = which(numcounts == 0)
  if(length(ind) > 0){
    X = X[, -ind]
    Y = Y[, -ind]
  }
  rm(numcounts)
  labelnames = as.character(unique(label))
  det_rate = matrix(NA, ncol(X), length(labelnames))
  gene_mean = matrix(NA, ncol(X), length(labelnames))
  
  for (i in 1:length(labelnames)){
    ind = which(label==labelnames[i])
    det_rate[,i] = colMeans(X[ind, ])
    gene_mean[,i] = apply(Y[ind, ], 2, get_positive_mean)
  }
  colnames(det_rate) = colnames(gene_mean) = labelnames

  dropout_rate = 1-det_rate
  rm(X, Y); gc()

  gene_mean = as.data.frame(gene_mean)
  det_rate = as.data.frame(det_rate)
  dropout_rate = as.data.frame(dropout_rate)

  gene_mean$id = paste0('gene', 1:nrow(gene_mean))
  det_rate$id = paste0('gene', 1:nrow(det_rate))
  dropout_rate$id = paste0('gene', 1:nrow(dropout_rate))

  mgm = melt(gene_mean)
  mdr = melt(det_rate)
  mdor = melt(dropout_rate)

  df = data.frame(gene = mgm$id,
                  celltype = mgm$variable,
                  det_rate = mdr$value,
                  normalized_mean = mgm$value,
                  dropout_rate = mdor$value)
  return(df)
}


preprocess_by_data = function(X){
  Y = X
  X = X>0 * 1

  numcounts = colSums(X)
  ind = which(numcounts == 0)
  if(length(ind) > 0){
    X = X[, -ind]
    Y = Y[, -ind]
  }
  rm(numcounts)
  det_rate = colMeans(X)
  gene_mean = apply(Y, 2, get_positive_mean)

  dropout_rate = 1-det_rate
  rm(X, Y); gc()

  gene_mean = as.data.frame(gene_mean)
  det_rate = as.data.frame(det_rate)
  dropout_rate = as.data.frame(dropout_rate)

  df = data.frame(det_rate = det_rate,
                normalized_mean = gene_mean,
                dropout_rate = dropout_rate)
  
  return(df)

}
```



```{r, echo = FALSE}
# X = readMM("../data/10X_pbmc_filtered/matrix.mtx")
# X = as.matrix(X)
# X = t(X)
# label1 = read.table("../data/10X_pbmc_filtered/truelabel.tsv")$V1
# levels(label1) = c("CD19+ B", "CD34+", "CD8+ Cytotoxic T", "CD4+ T Helper2", "CD4+/CD45RO+ Memory", "CD14+ Monocyte", "CD8+/CD45RA+ Naive Cytotoxic", "CD4+/CD45RA+/CD25- Naive T", "CD56+ NK", "CD4+/CD25 T Reg")
# label1_upper = c("CD19+", "CD34+", "CD8+", "CD4+", "CD4+", "CD14+", "CD8+", "CD4+", "CD56+", "CD4+")
# celltypemat = cbind(levels(label1), label1_upper)
# df = preprocess(X, label=label1, upper = label1_upper, normalize = FALSE)
# rm(X); gc()
# df$celltype2 = celltypemat[match(df$celltype, celltypemat[,1]),2]
# df$data = "purified_10X_pbmc"
# 
# X = t(as.matrix(readMM("../data/10X_68K/filtered_matrix/hg19/matrix.mtx")))
# annotation = read.table("../data/10X_68K/filtered_matrix/hg19/barocdes_annotation.tsv", sep = "\t", header = TRUE)
# label2 = annotation$celltype
# label2_upper = c("CD14+", "CD19+", "CD34+", "CD4+", "CD4+", "CD4+", "CD4+", "CD56+", "CD8+", "CD8+", "Dendritic")
# tmp = preprocess(X, label=label2, normalize = FALSE)
# tmp$data = "10X_68K"
# tmp$celltype2 = label2_upper[match(tmp$celltype, levels(label2))]
# df = rbind(df, tmp)
# print(dim(tmp))
# rm(tmp)
# rm(X); gc()
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc09_tumor1.rds")))
# label3 = read.table("../data/GSE114724/GSE114724_CLEANED/bc09_tumor1_label.txt")$V1
# levels(label3) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
# label3_upper = c("CD19+", "", "", "", "CD14+", "", "CD56+", "", "CD4+", "CD4+", "CD8+", "CD8+")
# tmp = preprocess(X, label=label3, normalize = FALSE)
# tmp$celltype2 = label3_upper[match(tmp$celltype, levels(label3))]
# tmp$data = "patient09tumor1"
# df = rbind(df, tmp)
# rm(X); gc()
# rm(tmp)
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc09_tumor2.rds")))
# label4 = read.table("../data/GSE114724/GSE114724_CLEANED/bc09_tumor2_label.txt")$V1
# levels(label4) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
# label4_upper = c("CD19+", "", "", "", "CD14+", "", "CD56+", "", "CD4+", "CD4+", "CD8+", "CD8+")
# tmp = preprocess(X, label=label4, normalize = FALSE)
# tmp$celltype2 = label4_upper[match(tmp$celltype, levels(label4))]
# tmp$data = "patient09tumor2"
# df = rbind(df, tmp)
# rm(X); gc()
# rm(tmp)
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc10_tumor1.rds")))
# label4 = read.table("../data/GSE114724/GSE114724_CLEANED/bc10_tumor1_label.txt")$V1
# levels(label4) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
# label4_upper = c("CD19+", "", "", "", "CD14+", "", "CD56+", "", "CD4+", "CD4+", "CD8+", "CD8+")
# tmp = preprocess(X, label=label4, normalize = FALSE)
# tmp$celltype2 = label4_upper[match(tmp$celltype, levels(label4))]
# tmp$data = "patient10tumor1"
# df = rbind(df, tmp)
# rm(X); gc()
# rm(tmp)
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc11_tumor1.rds")))
# label4 = read.table("../data/GSE114724/GSE114724_CLEANED/bc11_tumor1_label.txt")$V1
# levels(label4) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
# label4_upper = c("CD19+", "", "", "", "CD14+", "", "CD56+", "", "CD4+", "CD4+", "CD8+", "CD8+")
# tmp = preprocess(X, label=label4, normalize = FALSE)
# tmp$celltype2 = label4_upper[match(tmp$celltype, levels(label4))]
# tmp$data = "patient11tumor1"
# df = rbind(df, tmp)
# rm(X); gc()
# rm(tmp)
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc11_tumor2.rds")))
# label4 = read.table("../data/GSE114724/GSE114724_CLEANED/bc11_tumor2_label.txt")$V1
# levels(label4) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
# label4_upper = c("CD19+", "", "", "", "CD14+", "", "CD56+", "", "CD4+", "CD4+", "CD8+", "CD8+")
# tmp = preprocess(X, label=label4, normalize = FALSE)
# tmp$celltype2 = label4_upper[match(tmp$celltype, levels(label4))]
# tmp$data = "patient11tumor2"
# df = rbind(df, tmp)
# rm(X); gc()
# rm(tmp)
# 
# write.table(df, "../data/10X_combined_datasets_by_celltype.txt", col.names=TRUE, row.names=FALSE)

df = read.table("../data/10X_combined_datasets_by_celltype.txt", header=TRUE)
df = as_tibble(df)
```



# Introduction to data sets

The Purified PBMC cells and PBMC 68K cells come from the 10X website. The different tumors from different patients are from the paper "Single-Cell Map of Diverse Immune Phenotypes in the Breast Tumor Microenvironment" from Azizi et al. The labels except the purified PBMC cells are all taken from the original papers. (https://www.sciencedirect.com/science/article/pii/S0092867418307232?via%3Dihub, https://www.nature.com/articles/ncomms14049)

```{r, echo = FALSE, comment=NA}
label1 = read.table("../data/10X_pbmc_filtered/truelabel.tsv")$V1
levels(label1) = c("CD19+ B", "CD34+", "CD8+ Cytotoxic T", "CD4+ T Helper2", "CD4+/CD45RO+ Memory", "CD14+ Monocyte", "CD8+/CD45RA+ Naive Cytotoxic", "CD4+/CD45RA+/CD25- Naive T", "CD56+ NK", "CD4+/CD25 T Reg")
df1 = as.data.frame(table(label1)); colnames(df1) = c("label", "purified_10X_pbmc")

annotation = read.table("../data/10X_68K/filtered_matrix/hg19/barocdes_annotation.tsv", sep = "\t", header = TRUE)
label2 = annotation$celltype
df2 = as.data.frame(table(label2)); colnames(df2) = c("label", "10X_68K")

label3 = read.table("../data/GSE114724/GSE114724_CLEANED/bc09_tumor1_label.txt")$V1
levels(label3) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
df3 = as.data.frame(table(label3)); colnames(df3) = c("label", "patient09tumor1")

label4 = read.table("../data/GSE114724/GSE114724_CLEANED/bc09_tumor2_label.txt")$V1
levels(label4) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
df4 = as.data.frame(table(label4)); colnames(df4) = c("label", "patient09tumor2")

label5 = read.table("../data/GSE114724/GSE114724_CLEANED/bc10_tumor1_label.txt")$V1
levels(label5) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
df5 = as.data.frame(table(label5)); colnames(df5) = c("label", "patient10tumor1")

label6 = read.table("../data/GSE114724/GSE114724_CLEANED/bc11_tumor1_label.txt")$V1
levels(label6) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
df6 = as.data.frame(table(label6)); colnames(df6) = c("label", "patient11tumor1")

label7 = read.table("../data/GSE114724/GSE114724_CLEANED/bc11_tumor2_label.txt")$V1
levels(label7) = c("CD19+ B", "Macrophage", "Mast", "mDC", "CD14+ Monocyte", "Neutrophil", "CD56+ NK", "CD56+ NK", "NKT", "CD4+ CM","CD4+/CD45RA+/CD25- Naive T", "CD8+ EM",  "CD8+/CD45RA+ Naive Cytotoxic")
df7 = as.data.frame(table(label7)); colnames(df7) = c("label", "patient11tumor2")

labelcounts = full_join(full_join(full_join(full_join(full_join(full_join(df1, df2), df3), 
          df4), df5), df6), df7, by = "label")
labelcounts[is.na(labelcounts)] = 0
colnames(labelcounts) = c("Cell Type", "PBMC Purified", "PBMC 68K", "P9T1", "P9T2", "P10T1", "P11T1", "P11T2")
library(janitor)
labelcounts = labelcounts %>% adorn_totals("row")
print(labelcounts)
```

# By each data set

For a given data set, we inspected if different cell types share the same relationship between the drop out rates and the mean expression level. 

```{r, echo = FALSE}
summarize_by_datasets = function(dataname){
  tmpdata = df %>% filter(data==dataname) %>% 
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=99))) %>%
  group_by(range, celltype) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

fitdf = tmpdata %>% group_by(celltype) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
typenames = unique(tmpdata$celltype)

for (t in 1:length(typenames)){
  ind = which(tmpdata$celltype==typenames[t])
  tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
}

ggplot(tmpdata, aes(x=mean, y=mean_dropout_rate, col=celltype)) + 
  geom_point(alpha = 0.5) + geom_line(aes(x=mean, y=fitted, col=celltype), lwd=2, alpha = 0.3) +
  ylab("dropout rate") +
  xlab("gene mean") +
  ylim(c(0,1)) + 
  ggtitle(dataname)
}
```

```{r, cache=TRUE}
summarize_by_datasets("purified_10X_pbmc")
summarize_by_datasets("10X_68K")
summarize_by_datasets("patient09tumor1")
summarize_by_datasets("patient09tumor2")
summarize_by_datasets("patient10tumor1")
summarize_by_datasets("patient11tumor1")
summarize_by_datasets("patient11tumor2")
```


# By each cell type

Reversely, we fixed each cell type and compared across the data sets. 

```{r, echo=FALSE}
summarize_by_celltype = function(df, celltypename){
  tmpdata = df %>% filter(celltype==celltypename) %>% 
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=99))) %>%
  group_by(range, data) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

  datanames = names(table(tmpdata$data))[table(tmpdata$data)>0]
  fitdf = tmpdata %>% group_by(data) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
  decayfit_tidy = broom::tidy(fitdf, decayfit)
  t = 1
  for (t in 1:length(datanames)){
    ind = which(tmpdata$data==datanames[t])
    tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
  }

ggplot(tmpdata, aes(x=mean, y=mean_dropout_rate, col=data)) + 
  geom_point(alpha = 0.5) + geom_line(aes(x=mean, y=fitted, col=data), lwd=2, alpha = 0.3) +
  ylab("dropout rate") +
  xlab("gene mean") +
  ylim(c(0,1)) + 
  ggtitle(celltypename) +
  xlim(c(0.5,7))
}
```

```{r}
summarize_by_celltype(df, "CD34+")
summarize_by_celltype(df, "CD14+ Monocyte")
summarize_by_celltype(df, "CD19+ B")
summarize_by_celltype(df, "CD4+/CD45RA+/CD25- Naive T")
summarize_by_celltype(df, "CD56+ NK")
summarize_by_celltype(df, "CD8+/CD45RA+ Naive Cytotoxic")
```

# Verify that the curves are the same across cell types

First, CD14+ Monocyte.

```{r, echo = FALSE}
cellname = "CD14+ Monocyte"
df = read.table("../data/10X_combined_datasets_by_celltype.txt", header=TRUE)
df = as_tibble(df)
tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>%
  filter(celltype == cellname) %>%
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range, data) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
datanames = names(table(tmpdata$data))[table(tmpdata$data)>0]
fitdf = tmpdata %>% group_by(data) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
for (t in 1:length(datanames)){
  ind = which(tmpdata$data==datanames[t])
  tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
}

yf = decayfit_tidy %>% filter(term=="yf") %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
y0 = decayfit_tidy %>% filter(term=="y0")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
log_alpha = decayfit_tidy %>% filter(term=="log_alpha")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>% 
  filter(celltype == "CD14+ Monocyte") %>%
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
fitdf = tmpdata %>% do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
tmpdata$fitted = predict(fitdf$decayfit[1][[1]], newdata = tmpdata)

yfmax = decayfit_tidy$estimate + 2 * decayfit_tidy$std.error

tmp = data.frame(data = "combined", 
           term = decayfit_tidy$term,
           estimate = decayfit_tidy[,2],
           std.error = decayfit_tidy[,3],
           statistic = decayfit_tidy[,4],
           p.value = decayfit_tidy[,5])
tmp = tmp %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

yf = rbind(as.data.frame(yf), tmp[1,])
y0 = rbind(as.data.frame(y0), tmp[2,])
log_alpha = rbind(as.data.frame(log_alpha), tmp[3,])

ggplot(yf, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=yf[8, "min"], ymax=yf[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("yf")

ggplot(y0, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=y0[8, "min"], ymax=y0[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("y0")

ggplot(log_alpha, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=log_alpha[8, "min"], ymax=log_alpha[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("log_alpha")
```

Now, CD56+ NK.

```{r, echo = FALSE}
cellname = "CD56+ NK"
df = read.table("../data/10X_combined_datasets_by_celltype.txt", header=TRUE)
df = as_tibble(df)
tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>%
  filter(celltype == cellname) %>%
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range, data) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
datanames = names(table(tmpdata$data))[table(tmpdata$data)>0]
fitdf = tmpdata %>% group_by(data) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
for (t in 1:length(datanames)){
  ind = which(tmpdata$data==datanames[t])
  tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
}

yf = decayfit_tidy %>% filter(term=="yf") %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
y0 = decayfit_tidy %>% filter(term=="y0")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
log_alpha = decayfit_tidy %>% filter(term=="log_alpha")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>% 
  filter(celltype == "CD14+ Monocyte") %>%
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range) %>%
  summarize(mean = mean(normalized_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
fitdf = tmpdata %>% do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
tmpdata$fitted = predict(fitdf$decayfit[1][[1]], newdata = tmpdata)

yfmax = decayfit_tidy$estimate + 2 * decayfit_tidy$std.error

tmp = data.frame(data = "combined", 
           term = decayfit_tidy$term,
           estimate = decayfit_tidy[,2],
           std.error = decayfit_tidy[,3],
           statistic = decayfit_tidy[,4],
           p.value = decayfit_tidy[,5])
tmp = tmp %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

yf = rbind(as.data.frame(yf), tmp[1,])
y0 = rbind(as.data.frame(y0), tmp[2,])
log_alpha = rbind(as.data.frame(log_alpha), tmp[3,])

ggplot(yf, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=yf[8, "min"], ymax=yf[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("yf")

ggplot(y0, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=y0[8, "min"], ymax=y0[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("y0")

ggplot(log_alpha, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=log_alpha[8, "min"], ymax=log_alpha[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("log_alpha")
```


# By each data set across cell types

Regardless of cell type, we infer the curve for each data set, and compare if the curve changes across different data sets. We conclude that the behavior of CD34+ cells in 10X 68K data set is not universal, and so we removed them.


```{r, echo=FALSE}
# X = readMM("../data/10X_pbmc_filtered/matrix.mtx")
# X = as.matrix(X)
# X = t(X)
# df_purified = preprocess_by_data(X)
# df_purified$data = "purified_10X_pbmc"
# 
# X = t(as.matrix(readMM("../data/10X_68K/filtered_matrix/hg19/matrix.mtx")))
# annotation = read.table("../data/10X_68K/filtered_matrix/hg19/barocdes_annotation.tsv", sep = "\t", header = TRUE)
# label2 = annotation$celltype
# ind = which(label2 != "CD34+")
# X = X[ind,]
# df_68k = preprocess_by_data(X)
# df_68k$data = "10X_68K"
# rm(X); gc()
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc09_tumor1.rds")))
# df_tumor = preprocess_by_data(X)
# df_tumor$data = "patiemt09tumor1"
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc09_tumor2.rds")))
# df_tumor2 = preprocess_by_data(X)
# df_tumor2$data = "patiemt09tumor2"
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc10_tumor1.rds")))
# df_tumor3 = preprocess_by_data(X)
# df_tumor3$data = "patiemt10tumor1"
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc11_tumor1.rds")))
# df_tumor4 = preprocess_by_data(X)
# df_tumor4$data = "patiemt11tumor1"
# 
# X = t(as.matrix(readRDS("../data/GSE114724/GSE114724_CLEANED/bc11_tumor2.rds")))
# df_tumor5 = preprocess_by_data(X)
# df_tumor5$data = "patiemt11tumor2"
# 
# df = rbind(df_purified, df_68k, df_tumor, df_tumor2, df_tumor3, df_tumor4, df_tumor5)
# rm(df_purified, df_68k, df_tumor)
# 
# write.table(df, "../data/10X_combined_datasets_by_data.txt", col.names=TRUE, row.names=FALSE)

df = read.table("../data/10X_combined_datasets_by_data.txt", header=TRUE)
df = as_tibble(df)

```

```{r, echo = FALSE}
tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>% 
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range, data) %>%
  summarize(mean = mean(gene_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)


tmpdata$fitted = as.numeric(tmpdata$fitted)
datanames = names(table(tmpdata$data))[table(tmpdata$data)>0]
fitdf = tmpdata %>% group_by(data) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
for (t in 1:length(datanames)){
  ind = which(tmpdata$data==datanames[t])
  tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
}


ggplot(tmpdata, aes(x=mean, y=mean_dropout_rate, col=data)) + 
  geom_point(alpha = 0.5) + geom_line(aes(x=mean, y=fitted, col=data), lwd=2, alpha = 0.3) +
  ylab("dropout rate") +
  xlab("gene mean") +
  ylim(c(0,1)) + 
  ggtitle("") +
  xlim(c(0.5,7))
```


# Verify that the curves are the same across the data sets

Exponential decay involves three parameters. Once the curve is fitted, the R function returns both estimate and standard error for each parameter for each curve. Here we show that the confidence intervals across all 7 sets have plenty of overlapping with the confidence interval acquired from combined dataset. 

```{r, echo = FALSE}
df = read.table("../data/10X_combined_datasets_by_data.txt", header=TRUE)
df = as_tibble(df)
tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>%
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range, data) %>%
  summarize(mean = mean(gene_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
datanames = names(table(tmpdata$data))[table(tmpdata$data)>0]
fitdf = tmpdata %>% group_by(data) %>%  do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
for (t in 1:length(datanames)){
  ind = which(tmpdata$data==datanames[t])
  tmpdata$fitted[ind] = predict(fitdf$decayfit[t][[1]], newdata = tmpdata[ind,-5])
}

yf = decayfit_tidy %>% filter(term=="yf") %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
y0 = decayfit_tidy %>% filter(term=="y0")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)
log_alpha = decayfit_tidy %>% filter(term=="log_alpha")%>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

tmpdata = df %>% filter(dropout_rate <= 0.99 & dropout_rate >= 0.01) %>% 
  mutate(range = cut(dropout_rate, breaks = seq(0.01,0.99,length=49))) %>%
  group_by(range) %>%
  summarize(mean = mean(gene_mean)) %>%
  mutate(minrange = substring(sapply(strsplit(as.character(range), ","), function(x) x[1]), 2)) %>%
  mutate(maxrange = gsub('.{1}$', '', sapply(strsplit(as.character(range), ','), function(x) x[2]) )) %>%
  mutate(minrange = as.numeric(minrange)) %>%
  mutate(maxrange = as.numeric(maxrange)) %>%
  mutate(mean_dropout_rate = (minrange + maxrange)/2) %>%
  select(-range) %>% select(-minrange) %>% select(-maxrange) %>%
  mutate(fitted = NA)

tmpdata$fitted = as.numeric(tmpdata$fitted)
fitdf = tmpdata %>% do(decayfit = nls(mean_dropout_rate ~ SSasymp(mean, yf, y0, log_alpha), data=.))
decayfit_tidy = broom::tidy(fitdf, decayfit)
tmpdata$fitted = predict(fitdf$decayfit[1][[1]], newdata = tmpdata)

yfmax = decayfit_tidy$estimate + 2 * decayfit_tidy$std.error

tmp = data.frame(data = "combined", 
           term = decayfit_tidy$term,
           estimate = decayfit_tidy[,2],
           std.error = decayfit_tidy[,3],
           statistic = decayfit_tidy[,4],
           p.value = decayfit_tidy[,5])
tmp = tmp %>%
  mutate(max = estimate + 2 * std.error) %>%
  mutate(min = estimate - 2 * std.error)

yf = rbind(as.data.frame(yf), tmp[1,])
y0 = rbind(as.data.frame(y0), tmp[2,])
log_alpha = rbind(as.data.frame(log_alpha), tmp[3,])



ggplot(yf, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=yf[8, "min"], ymax=yf[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("yf")

ggplot(y0, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=y0[8, "min"], ymax=y0[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("y0")

ggplot(log_alpha, aes(x = data)) +
  geom_errorbar(aes(ymax = max, ymin = min)) +
  geom_point(aes(y=estimate)) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=log_alpha[8, "min"], ymax=log_alpha[8, "max"]), alpha = 0.02) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("log_alpha")
```